---
title: "Gene module analysis"
author: "Mikayla, Mustafa, Yibo"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

This report is totally original without taking code from the github repo of the original paper, (mostly because we find their code is too hard to read and relied on obsolete packages). This should also include the output of each section but was accidentally deleted. We couldn’t reproduce the output because it took too long to re-run the whole code that we might miss the deadline.

# The first part of the code did all the data reading, normalizing, PCA, and clusetring at different resolution.

```{r}
library(Seurat)
library(Matrix)

# List of patient IDs
patient_ids <- c("CID3586", "CID3838", "CID3921", "CID3941", "CID3946", "CID3948", "CID3963", "CID4040", "CID4066", "CID4067", "CID4290A", "CID4398", "CID4461", "CID4463", "CID4465", "CID4471", "CID4495", "CID4513", "CID4515", "CID4523", "CID4530N", "CID4535", "CID44041", "CID44971", "CID44991", "CID45171")

# Initialize an empty list to store the Seurat objects
seurat_objects <- list()

# Loop through each patient
for (patient_id in patient_ids) {
  # Define file paths for the patient's data
  count_matrix_barcodes_path <- paste0("./GSE176078_RAW/", patient_id, "/count_matrix_barcodes.tsv")
  count_matrix_genes_path <- paste0("./GSE176078_RAW/", patient_id, "/count_matrix_genes.tsv")
  count_matrix_sparse_path <- paste0("./GSE176078_RAW/", patient_id, "/count_matrix_sparse.mtx")
  metadata_path <- paste0("./GSE176078_RAW/", patient_id, "/metadata.csv")
  
  # Load count matrix and metadata
  count_matrix <- readMM(file = count_matrix_sparse_path)
  barcodes <- read.delim(count_matrix_barcodes_path, header = FALSE, col.names = c("barcode"))
  genes <- read.delim(count_matrix_genes_path, header = FALSE, col.names = c("gene"))
  metadata <- read.csv(metadata_path, row.names = 1)
  
  # Filter cells based on the celltype_major column
  cells_to_keep <- metadata$celltype_major %in% c("Cancer Epithelial", "CAFs")
  
  # Set the colnames and rownames of the count matrix
  colnames(count_matrix) <- barcodes$barcode
  rownames(count_matrix) <- genes$gene
  
  # Subset the count matrix using the cells_to_keep boolean vector
  count_matrix <- count_matrix[, cells_to_keep]
  
  # Subset the metadata using cells_to_keep
  metadata <- metadata[cells_to_keep, , drop = FALSE]
  
  # Create Seurat object
  seurat_obj <- CreateSeuratObject(counts = count_matrix, project = patient_id, assay = "RNA", meta.data = metadata)
  
  # Preprocessing, normalization, and scaling
  seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = 10000)
  seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)
  seurat_obj <- ScaleData(seurat_obj, features = rownames(seurat_obj))
  
  # Run PCA on the scaled data
  seurat_obj <- RunPCA(seurat_obj, features = VariableFeatures(object = seurat_obj))
  
  # Create a shared nearest neighbor graph
  seurat_obj <- FindNeighbors(seurat_obj, reduction = "pca", dims = 1:20)
  
  # Find clusters at different resolutions
  resolutions <- c(0.4, 0.8, 1.2, 1.6, 2.0)
  for (res in resolutions) {
    seurat_obj <- FindClusters(seurat_obj, resolution = res, ident.1 = "neoplastic", algorithm = 3)
  }
  
  # Add the processed Seurat object to the list
  seurat_objects[[patient_id]] <- seurat_obj
}

# Print the list of Seurat objects to check the results
seurat_objects
```
# Identify all DEGs
```{r}
# Initialize an empty list to store the differentially expressed genes (DEGs) for each patient at different resolutions
patient_DEGs <- list()

# Loop through each patient
for (patient_id in patient_ids) {
  # Get the Seurat object for the patient
  seurat_obj <- seurat_objects[[patient_id]]

  # Initialize an empty list to store the DEGs for each resolution
  resolution_DEGs <- list()

  # Loop through each resolution
  for (res in resolutions) {
    # Update the active identity to the one corresponding to the current resolution
    Idents(seurat_obj) <- paste0("RNA_snn_res.", res)

    # Find the top 200 differentially expressed genes for each cluster
    DEGs <- FindAllMarkers(seurat_obj, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, test.use = "wilcox", max.cells.per.ident = Inf, min.cells.group = 5, max.features = 200)

    # Add the DEGs to the list
    resolution_DEGs[[as.character(res)]] <- DEGs
  }

  # Add the DEGs at different resolutions to the list
  patient_DEGs[[patient_id]] <- resolution_DEGs
}
```

# Check data structure
```{r}
#str(patient_DEGs)
```

# filter to preserve only top 200 DEGs
```{r}
# Initialize an empty list to store the differentially expressed genes (DEGs) for each patient at different resolutions
patient_DEGs_top200 <- list()

# Loop through each patient
for (patient_id in patient_ids) {
  # Get the DEGs for the patient at different resolutions
  patient_res_DEGs <- patient_DEGs[[patient_id]]

  # Initialize an empty list to store the top 200 DEGs for each resolution
  resolution_DEGs_top200 <- list()

  # Loop through each resolution
  for (res in names(patient_res_DEGs)) {
    # Get the DEGs for the current resolution
    res_DEGs <- patient_res_DEGs[[res]]

    # Initialize an empty list to store the top 200 DEGs for each cluster
    cluster_DEGs_top200 <- list()

    # Get unique cluster IDs
    unique_clusters <- unique(res_DEGs$cluster)

    # Loop through each cluster
    for (cluster_id in unique_clusters) {
      # Filter the DEGs for the current cluster
      cluster_DEGs <- res_DEGs[res_DEGs$cluster == cluster_id, ]

      # Get the top 200 DEGs for the current cluster
      top200_DEGs <- head(cluster_DEGs, n = min(200, nrow(cluster_DEGs)))

      # Add the top 200 DEGs for the current cluster to the list
      cluster_DEGs_top200[[cluster_id]] <- top200_DEGs
    }

    # Add the top 200 DEGs for each cluster at the current resolution to the list
    resolution_DEGs_top200[[res]] <- cluster_DEGs_top200
  }

  # Add the top 200 DEGs at different resolutions for the current patient to the list
  patient_DEGs_top200[[patient_id]] <- resolution_DEGs_top200
}
```

# Check the result
```{r}
str(patient_DEGs_top200)
```

# filter to keep n.gene>5 and n.cell>5

```{r}
# Initialize an empty list to store the filtered DEGs for each patient at different resolutions
patient_DEGs_filtered <- list()

# Loop through each patient
for (patient_id in patient_ids) {
  cat("Patient:", patient_id, "\n")
  
  # Get the top 200 DEGs for the patient at different resolutions
  patient_res_DEGs_top200 <- patient_DEGs_top200[[patient_id]]

  # Get the metadata of the Seurat object for the current patient
  metadata <- seurat_objects[[patient_id]]@meta.data

  # Initialize an empty list to store the filtered DEGs for each resolution
  resolution_DEGs_filtered <- list()

  # Loop through each resolution
  for (res in names(patient_res_DEGs_top200)) {
    cat("  Resolution:", res, "\n")
    
    # Get the top 200 DEGs for the current resolution
    res_cluster_DEGs_top200 <- patient_res_DEGs_top200[[res]]

    # Get the cluster column name for the current resolution
    cluster_col_name <- paste0("RNA_snn_res.", res)

    # Calculate the number of cells in each cluster at the current resolution
    cluster_counts <- table(metadata[[cluster_col_name]])
    cat("    Cluster cell counts:", cluster_counts, "\n")

    # Initialize an empty list to store the filtered DEGs for each cluster
    cluster_DEGs_filtered <- list()

    # Loop through each cluster
    for (cluster_id in names(res_cluster_DEGs_top200)) {
      # Get the top 200 DEGs for the current cluster
      cluster_top200_DEGs <- res_cluster_DEGs_top200[[cluster_id]]

      # Get the number of cells in the current cluster
      cluster_cell_count <- cluster_counts[[cluster_id]]

      # Check if the gene set contains more than 5 genes and originates from a cluster with more than 5 cells
      if (nrow(cluster_top200_DEGs) > 5 && cluster_cell_count > 5) {
        # Add the filtered DEGs for the current cluster to the list
        cluster_DEGs_filtered[[cluster_id]] <- cluster_top200_DEGs
      }
    }

    # Add the filtered DEGs for each cluster at the current resolution to the list
    resolution_DEGs_filtered[[res]] <- cluster_DEGs_filtered
    cat("    Number of filtered clusters:", length(cluster_DEGs_filtered), "\n")
  }

  # Add the filtered DEGs at different resolutions for the current patient to the list
  patient_DEGs_filtered[[patient_id]] <- resolution_DEGs_filtered
}
```

# Compare the filtered result with the original result to ensure the filter process is okay
```{r}
# Count the number of clusters in patient_DEGs_filtered
total_clusters_filtered <- sum(sapply(patient_DEGs_filtered, function(x) sum(sapply(x, length))))

# Count the number of clusters in patient_DEGs_top200
total_clusters_top200 <- sum(sapply(patient_DEGs_top200, function(x) sum(sapply(x, length))))

cat("Total number of clusters in patient_DEGs_filtered:", total_clusters_filtered, "\n")
cat("Total number of clusters in patient_DEGs_top200:", total_clusters_top200, "\n")
```
# remove redundancy within one sample according to jaccard index > 0.75
```{r}
for (patient_id in patient_ids) {
  patient_res_DEGs_filtered <- patient_DEGs_filtered[[patient_id]]
  
  # Loop through each resolution
  for (res in names(patient_res_DEGs_filtered)) {
    res_cluster_DEGs_filtered <- patient_res_DEGs_filtered[[res]]

    # Check if the res_cluster_DEGs_filtered is not empty
    if (length(res_cluster_DEGs_filtered) > 0) {
      # Calculate Jaccard index for all pairs of gene sets
      jaccard_values <- matrix(0, nrow = length(res_cluster_DEGs_filtered), ncol = length(res_cluster_DEGs_filtered))
      for (i in 1:(length(res_cluster_DEGs_filtered) - 1)) {
        for (j in (i + 1):length(res_cluster_DEGs_filtered)) {
          set1 <- rownames(res_cluster_DEGs_filtered[[i]])
          set2 <- rownames(res_cluster_DEGs_filtered[[j]])
          jaccard_values[i, j] <- jaccard(set1, set2)
        }
      }

      cat("Number of gene set pairs with Jaccard index greater than 0.75 in patient", patient_id, "resolution", res, ":", sum(jaccard_values > 0.75), "\n")

      # Find pairs with Jaccard index greater than 0.75 and remove the one with fewer genes
      to_remove <- c()
      for (i in 1:(length(res_cluster_DEGs_filtered) - 1)) {
        for (j in (i + 1):length(res_cluster_DEGs_filtered)) {
          if (jaccard_values[i, j] > 0.75) {
            if (length(rownames(res_cluster_DEGs_filtered[[i]])) < length(rownames(res_cluster_DEGs_filtered[[j]]))) {
              to_remove <- c(to_remove, i)
            } else {
              to_remove <- c(to_remove, j)
            }
          }
        }
      }

      cat("Number of gene sets removed in patient", patient_id, "resolution", res, ":", length(to_remove), "\n")

      # Remove the gene sets identified above if to_remove is not empty
      if (length(to_remove) > 0) {
        res_cluster_DEGs_filtered <- res_cluster_DEGs_filtered[-to_remove]
      }

      # Update patient_DEGs_filtered
      patient_res_DEGs_filtered[[res]] <- res_cluster_DEGs_filtered
    } else {
      cat("No clusters found in patient", patient_id, "resolution", res, "\n")
    }
  }

  # Update patient_DEGs_filtered
  patient_DEGs_filtered[[patient_id]] <- patient_res_DEGs_filtered
}
```

# Identify gene modules
```{r}
library(cola)

# Combine all gene signatures from patient_DEGs_filtered
all_gene_signatures <- list()
for (patient_id in patient_ids) {
  for (res in names(patient_DEGs_filtered[[patient_id]])) {
    all_gene_signatures <- c(all_gene_signatures, patient_DEGs_filtered[[patient_id]][[res]])
  }
}

# Calculate Jaccard similarity matrix
jaccard_similarity_matrix <- matrix(0, nrow = length(all_gene_signatures), ncol = length(all_gene_signatures))
for (i in 1:(length(all_gene_signatures) - 1)) {
  for (j in (i + 1):length(all_gene_signatures)) {
    set1 <- rownames(all_gene_signatures[[i]])
    set2 <- rownames(all_gene_signatures[[j]])
    jaccard_similarity_matrix[i, j] <- jaccard(set1, set2)
    jaccard_similarity_matrix[j, i] <- jaccard_similarity_matrix[i, j]
  }
}
```

# Find the optimal K value, top_value_method and parition_method
```{r}
library(cola)

r1 = run_all_consensus_partition_methods(jaccard_similarity_matrix, top_value_method = c("SD","MAD"), partition_method = c("skmeans", "kmeans"), max_k = 10)

cola_report(r1, output_dir = "C:/yc/cb_project/scRNASeq_analysis/")
```


#Use the matrix above to do consensus_clustering using skmeans
```{r}
# Perform consensus clustering using skmeans from cola package
set.seed(42) # Set seed for reproducibility
k <- 6 # Number of gene modules
consensus_clustering <- consensus_partition(jaccard_similarity_matrix, partition_method = "skmeans", top_value_method = "SD", max_k = k)
```

# Find the 6 gene modules
```{r}
library(dplyr)

# Get the class labels for each gene signature
class_labels <- consensus_clustering@object_list[["6"]][["class_df"]][["class"]]

# Assign class labels to each gene signature in all_gene_signatures
labeled_gene_signatures <- lapply(seq_along(all_gene_signatures), function(i) {
  gene_signature <- all_gene_signatures[[i]]
  gene_signature$module <- class_labels[i]
  return(gene_signature)
})

# Combine all labeled gene signatures into one data frame
combined_gene_signatures <- do.call(rbind, labeled_gene_signatures)

# Identify the top 200 most commonly expressed genes for each clustered gene module
top_200_genes_per_module <- combined_gene_signatures %>%
  group_by(module, gene) %>%
  summarise(count = n()) %>%
  arrange(module, desc(count)) %>%
  group_by(module) %>%
  slice_max(count, n = 200)

# Split the top 200 genes per module into a list of data frames
top_200_genes_list <- split(top_200_genes_per_module, top_200_genes_per_module$module)
```

# Calculate gene module scores based on the module above and AUCell
```{r}
library(Matrix)
library(AUCell)

patient_ids <- c("CID3586", "CID3838", "CID3921", "CID3941", "CID3946", "CID3948", "CID3963", "CID4040", "CID4066", "CID4067", "CID4290A", "CID4398", "CID4461", "CID4463", "CID4465", "CID4471", "CID4495", "CID4513", "CID4515", "CID4523", "CID4530N", "CID4535", "CID44041", "CID44971", "CID44991", "CID45171")

aucell_scores_all_patients <- list()

for (i in 1:length(patient_ids)) {
  patient_id <- patient_ids[i]
  cat("Dealing with", i, "of", length(patient_ids), "patients:", patient_id, "\n")

  count_matrix_barcodes_path <- paste0("./GSE176078_RAW/", patient_id, "/count_matrix_barcodes.tsv")
  count_matrix_genes_path <- paste0("./GSE176078_RAW/", patient_id, "/count_matrix_genes.tsv")
  count_matrix_sparse_path <- paste0("./GSE176078_RAW/", patient_id, "/count_matrix_sparse.mtx")
   metadata_path <- paste0("./GSE176078_RAW/", patient_id, "/metadata.csv")
  
  count_matrix <- readMM(file = count_matrix_sparse_path)
  barcodes <- read.delim(count_matrix_barcodes_path, header = FALSE, col.names = c("barcode"))
  genes <- read.delim(count_matrix_genes_path, header = FALSE, col.names = c("gene"))
  metadata <- read.csv(metadata_path, row.names = 1)
  
  # Filter cells based on the celltype_major column
  cells_to_keep <- metadata$celltype_major %in% c("Cancer Epithelial", "CAFs")
  # Get the indices of the cells to keep
  cells_to_keep_indices <- which(cells_to_keep)
  
  # Set the colnames and rownames of the count matrix
  colnames(count_matrix) <- barcodes$barcode
  rownames(count_matrix) <- genes$gene

  # Subset the count matrix using the cells_to_keep_indices
  count_matrix <- count_matrix[, cells_to_keep_indices]
  
  exprMatrix <- as.matrix(count_matrix)  # Convert the dgTMatrix to a regular matrix
  
  # Calculate enrichment scores for the top 200 genes of each module
  aucell_scores_list <- list()
  for (j in 1:length(top_200_genes_list)) {
    geneSet <- top_200_genes_list[[j]]$gene
    aucell_scores <- AUCell_run(exprMatrix, list(geneSet=geneSet), aucMaxRank=nrow(exprMatrix)*0.05)
    aucell_scores_list[[paste0("GM", j)]] <- aucell_scores
    
    # Calculate percentage of cells processed
    percentage_processed <- (j / length(top_200_genes_list)) * 100
    cat("Processed", round(percentage_processed, 2), "% cells of patient", patient_id, "\n")
  }
  
  aucell_scores_all_patients[[patient_id]] <- aucell_scores_list
}

# Print the list of AUCell scores for each patient
aucell_scores_all_patients
```

# Assign a gene module to each cell based on the highest AUCell score
```{r}
library(SummarizedExperiment)

assign_gene_module <- function(patient_id) {
  cell_ids <- colnames(aucell_scores_all_patients[[patient_id]][["GM1"]])
  
  cell_module_assignments <- sapply(cell_ids, function(cell_id) {
    # Get the scores for the selected cell in each gene module
    cell_scores <- sapply(names(aucell_scores_all_patients[[patient_id]]), function(module) {
      as.numeric(assays(aucell_scores_all_patients[[patient_id]][[module]])$AUC[, cell_id])
    })

    # Find the gene module with the highest score
    max_module <- names(cell_scores)[which.max(unlist(cell_scores))]
    
    return(max_module)
  }, simplify = FALSE)
  
  names(cell_module_assignments) <- cell_ids
  return(cell_module_assignments)
}

cell_gene_module_assignments <- lapply(names(aucell_scores_all_patients), assign_gene_module)
names(cell_gene_module_assignments) <- names(aucell_scores_all_patients)
```


# Calculate gini simpson index and entropy for each patient.
```{r}
# Function to calculate Gini-Simpson index
gini_simpson <- function(counts) {
  total_counts <- sum(counts)
  proportions <- counts / total_counts
  return(1 - sum(proportions^2))
}

# Function to calculate entropy
entropy <- function(counts) {
  total_counts <- sum(counts)
  proportions <- counts / total_counts
  return(-sum(proportions * log2(proportions)))
}

# Function to calculate coefficient of variation (CV)
cv <- function(counts) {
  total_counts <- sum(counts)
  proportions <- counts / total_counts
  mean_proportions <- mean(proportions)
  sd_proportions <- sd(proportions)
  return(sd_proportions / mean_proportions)
}

# Calculate Gini-Simpson index, entropy, and CV for each patient
gini_simpson_list <- list()
entropy_list <- list()
cv_list <- list()

for (patient_id in names(cell_gene_module_assignments)) {
  gene_module_assignments <- unlist(cell_gene_module_assignments[[patient_id]])
  gene_module_counts <- table(gene_module_assignments)
  gini_simpson_list[[patient_id]] <- gini_simpson(gene_module_counts)
  entropy_list[[patient_id]] <- entropy(gene_module_counts)
  cv_list[[patient_id]] <- cv(gene_module_counts)
}

# Combine results into a data frame
result <- data.frame(
  Patient_ID = names(cell_gene_module_assignments),
  Gini_Simpson = unlist(gini_simpson_list),
  Entropy = unlist(entropy_list),
  CV = unlist(cv_list)
)

print(result)

write.csv(result, file = "gene_module_ok.csv", row.names = FALSE)
```


# Visualize the calculated entropy, CV and GS index.
```{r}
library(ggplot2)
library(reshape2)

# Replace Patient_ID with numbers 1 to 26
result$Patient_ID <- as.factor(seq(1, nrow(result)))

# Melt the data frame for ggplot
melted_result <- melt(result, id.vars = "Patient_ID", variable.name = "Metric", value.name = "Value")

# Create the bar plot
p <- ggplot(melted_result, aes(x = Patient_ID, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
  labs(x = "Patient Number", y = "Value", title = "Metrics for Patients") +
  facet_wrap(~Metric, scales = "free_y")

print(p)
```


